Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2024-01-22 main.tex
:pencil: Add sidenote
:construction: Update Pareto singular front
:pencil: Update abstract

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{bcd}{BCD}{Block Coordinate Descent}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{siso}{SISO}{Single-Input Single-Output}
\newacronym{miso}{MISO}{Multiple-Input Single-Output}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}
\newacronym{wsr}{WSR}{Weighted Sum-Rate}
\newacronym{svd}{SVD}{Singular Value Decomposition}
\newacronym{mmse}{MMSE}{Minimum Mean-Square Error}
\newacronym{wmmse}{WMMSE}{Weighted \gls{mmse}}
\newacronym{mse}{MSE}{Mean-Square Error}

@@

\begin{abstract}
	\gls{ris} recycles and redistributes ambient waves for improved wireless performance.
	% \gls{ris} is a promising technology that recycles and redistributes ambient waves for improved wireless performance.
	% \gls{ris} is a real-time programmable reflector that recycles and redistributes ambient waves for improved wireless performance.
	This paper investigates the use of passive \gls{ris} for channel shaping in \gls{mimo} \gls{pc} and \gls{ic}.
	We depart from the widely-adapted diagonal phase shift model and focus on a general asymmetric \gls{bd} architecture, which allows off-diagonal entries by in-group connections between elements.
	For arbitrary group size, we propose a closed-form iterative algorithm for quadratic problems and a \gls{rcg} algorithm for general problems.
	Various case studies are then conducted, including Pareto singular value characterization and rate maximization for \gls{pc}, and interference alignment and \gls{wsr} maximization for \gls{ic}.
	To quantify channel shaping capability, we also derive analytical bounds under specific scenarios and validate those by aforementioned methods.
	Simulation results suggest the advantage of \gls{bd} \gls{ris} scales with group size, while dyadic or tetradic architecture usually strikes a good balance between performance and complexity.
	% Through theoretical analysis and numerical evaluation,
	Our findings unlock unprecedented efficiency and adaptability, paving the way for smarter and greener wireless environments.
	% truly programmable
	% useful guidelines for prototype design.
\end{abstract}

\begin{IEEEkeywords}
	Reconfigurable intelligent surface, channel shaping, point-to-point channel, interference channel.
\end{IEEEkeywords}

@@

\begin{section}{Introduction}
	% The quest for reliable, high-speed, and ubiquitous wireless connectivity has been long-standing since Marconi's illuminating radio in 1895.
	% Great successes have been made at transmitter and receiver sides over the past century, and the society is unprecedentedly close to the Shannon limit \cite{Shannon1948}.
	Today we are witnessing a paradigm shift from {connectivity} to {intelligence}, where the wireless environment is no longer a chaotic medium but a conscious agent that serves on demand.
	This is made possible by the recent advances in \gls{ris}, a real-time programmable metasurface made of numerous non-resonant sub-wavelength scattering elements.
	It can manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019} with a higher energy efficiency, lower cost, lighter footprint, and greater scalability than conventional relays.
	Using \gls{ris} for \emph{passive beamforming} has attracted significant interest in wireless communication \cite{Wu2019,Wu2020c,Yang2020,Zheng2021}, backscatter \cite{Jia2020,Liang2022}, sensing \cite{Liu2022a,Hua2023}, and power transfer literature \cite{Wu2021d,Feng2022,Zhao2022}, reporting a second-order array gain and fourth-order power scaling law (with proper waveform).
	On the other hand, \gls{ris} also enables \emph{backscatter modulation} by dynamically switching between different patterns, as already investigated \cite{Karasik2020,Basar2020,Zhao2022a} and prototyped \cite{Tang2019a,Dai2020a}.
	Although fruitful outcomes have been harvested over optimization tools, one fundamental yet critical question is the \emph{channel shaping} capability: To what extent can a passive \gls{ris} reshape the wireless channel?
	The answer indeed depends on the hardware architecture and scattering model.

	In conventional (a.k.a. diagonal) \gls{ris}, each scattering element is tuned by a dedicated impedance and functions as an \emph{individual} phase shifter \cite{Wu2020}.
	% , accounting for a diagonal phase shift model with unit magnitude.
	The concept is generalized to \gls{bd} \gls{ris} \cite{Shen2020a,Li2023b} featuring in-group connections between elements.
	This allows \emph{cooperative} reflection --- wave impinging on one element can propagate within the circuit and depart partially from all elements.
	\gls{bd} \gls{ris} can thus control both amplitude and phase of the reflected wave, generalizing the scattering matrix from diagonal with unit-magnitude entries to block diagonal with  unitary blocks.
	Its benefit has been recently shown in receive power maximization \cite{Nerini2023,Santamaria2023,Fang2023,Nerini2023a}, transmit power minimization \cite{Zhou2023}, and rate maximization \cite{Zhou2023,Nerini2023a,Li2023d,Bartoli2023,Li2023c}.
	The channel estimation \cite{Li2023e} and mutual coupling \cite{Li2023f} issues have also been investigated.
	Therefore, \gls{bd} \gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

@@

>>>>

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{bcd}{BCD}{Block Coordinate Descent}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{siso}{SISO}{Single-Input Single-Output}
\newacronym{miso}{MISO}{Multiple-Input Single-Output}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}
\newacronym{wsr}{WSR}{Weighted Sum-Rate}
\newacronym{svd}{SVD}{Singular Value Decomposition}
\newacronym{mmse}{MMSE}{Minimum Mean-Square Error}
\newacronym{wmmse}{WMMSE}{Weighted \gls{mmse}}
\newacronym{mse}{MSE}{Mean-Square Error}
\newacronym{los}{LoS}{Line-of-Sight}

@@

\begin{abstract}
	This paper investigates how a passive \gls{ris} can shape the \gls{mimo} point-to-point channel in terms of singular values.
	We depart from the widely-adapted diagonal phase shift model to a general \gls{bd} architecture, which provides superior shaping capability thanks to in-group connections between elements.
	An efficient \gls{rcg} algorithm is tailored for smooth optimization problems of asymmetric \gls{bd} \gls{ris} with arbitrary group size, then invoked for the Pareto frontier of channel singular values.
	To understand the gain from off-diagonal entries, we also derive analytical singular value bounds in \gls{los} and fully-connected scenarios.
	As a side product, we tackle \gls{mimo} rate maximization problem by alternating between active beamformer (eigenmode transmission) and passive beamformer (\gls{rcg} algorithm) until convergence.
	A low-complexity suboptimal solution that decouples both blocks is also proposed, where the channel shaping subproblem is formulated as power maximization and solved iteratively in closed form.
	Theoretical analysis and numerical evaluation reveal that the shaping advantage of \gls{bd} \gls{ris} increases with group size and stems from subchannel rearrangement and subspace alignment.
\end{abstract}

\begin{IEEEkeywords}
	Reconfigurable intelligent surface, multi-input multi-output, manifold optimization, singular value control, rate maximization.
\end{IEEEkeywords}

@@

\begin{section}{Introduction}
	% The quest for reliable, high-speed, and ubiquitous wireless connectivity has been long-standing since Marconi's illuminating radio in 1895.
	% Great successes have been made at transmitter and receiver sides over the past century, and the society is unprecedentedly close to the Shannon limit \cite{Shannon1948}.
	Today we are witnessing a paradigm shift from connectivity to intelligence, where the wireless environment is no longer a chaotic medium but a conscious agent that serves on demand.
	This is empowered by the recent advances in \gls{ris}, a real-time programmable metasurface of numerous non-resonant sub-wavelength scattering elements.
	It can manipulate the amplitude, phase, frequency, and polarization of the scattered waves \cite{Basar2019} with a higher energy efficiency, lower cost, lighter footprint, and greater scalability than relays.
	Using \gls{ris} for {passive beamforming} has attracted significant interest in wireless communication \cite{Wu2019,Wu2020c,Yang2020,Zheng2021}, backscatter \cite{Jia2020,Liang2022}, sensing \cite{Liu2022a,Hua2023}, and power transfer literature \cite{Wu2021d,Feng2022,Zhao2022}, reporting a second-order array gain and fourth-order power scaling law (with proper waveform).
	On the other hand, \gls{ris} also enables {backscatter modulation} by dynamically switching between different patterns, as already investigated \cite{Karasik2020,Basar2020,Zhao2022a} and prototyped \cite{Tang2019a,Dai2020a}.
	Although fruitful outcomes have been harvested, one fundamental unanswered question is the {channel shaping} capability: \emph{To what extent can a passive \gls{ris} reshape the wireless channel?}

	The answer indeed depends on the hardware architecture and scattering model.
	In conventional (a.k.a. diagonal) \gls{ris}, each scattering element is tuned by a dedicated impedance and acts as an \emph{individual} phase shifter \cite{Wu2020}.
	% The concept is generalized to \gls{bd} \gls{ris} \cite{Shen2020a,Li2023b} featuring in-group connections between elements.
	The concept is generalized to \gls{bd} \gls{ris} \cite{Shen2020a,Li2023b} that connects adjacent elements as a group using passive components.
	This allows \emph{cooperative} scattering --- wave impinging on one element can propagate within the circuit and depart partially from any elements in the same group.
	\gls{bd} \gls{ris} can thus control both amplitude and phase of the reflected wave, generalizing the scattering matrix from diagonal with unit-magnitude entries to block diagonal with  unitary blocks.
	Its benefit has been recently shown in receive power maximization \cite{Nerini2023,Santamaria2023,Fang2023,Nerini2023a}, transmit power minimization \cite{Zhou2023}, and rate maximization \cite{Zhou2023,Nerini2023a,Li2023d,Bartoli2023,Li2023c}.
	The channel estimation \cite{Li2023e} and mutual coupling \cite{Li2023f} issues have also been investigated.
	Therefore, \gls{bd} \gls{ris} is envisioned as the next generation channel shaper with stronger signal processing flexibility \cite{Li2023g}.

@@

% ! Evaluate the rate performance of RIS that maximizes channel power (should coincide at low and high snr)
